---
title: "Visualisation using ggplot2 - part 1"
subtitle: "R cafe" 
author: "Anke Tukker"
date: "November 16, 2022"
output: 
  ioslides_presentation:
    widescreen: TRUE
    smaller: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Graphs in R

Multiple packages:

-   Basic R plot()
    -   Helpful for first data exploration
-   ggplot2
    -   Allows for many modifications

```{r}
library(ggplot2)
```

## Quick example {.smaller .columns-2}

[Diamonds dataset](https://ggplot2.tidyverse.org/reference/diamonds.html)

Base plots can be quick

```{r, out.width = '500px', include = TRUE, echo = TRUE, message = FALSE}
hist(diamonds$price)
```

But so is ggplot

```{r, out.width = '500px', include = TRUE, echo = TRUE, message = FALSE}
ggplot(diamonds)+
  geom_histogram(aes(x = price))
```

```{r, out.width = '500px', include = TRUE, echo = TRUE, message = FALSE}
#Try using "carat", "x", "y", "z", "depth" as an input for x
ggplot(diamonds)+
  geom_histogram(aes(x = ))

```

##  {.smaller .columns-2}

Making the plot more aesthetically appealing

```{r, out.width = '500px', include = TRUE, echo = TRUE, message = FALSE}
ggplot(diamonds, aes(price, fill = cut)) +
  geom_histogram()
```

```{r, out.width = '500px', include = TRUE, echo = TRUE, message = FALSE}
# fill the plot with "color", or "clarity"
ggplot(diamonds, aes(price, fill =  )) +
  geom_histogram()
```

## What is ggplot2 and why use it?

A package that is

-   elegant and versatile
-   grammar of graphics
-   layered structure that lets you build block by block
-   different from base graphics (base R)
    -   curve could be steep
        -   but looks amazing when it works!
-   more information

## Prepare library

Options:

-   tidyverse
    -   includes many packages (ggplot2, dplyr)
    -   time to load
-   ggplot2 <!-- \ means line break -->\

```{r, echo = TRUE, results = 'hide', error = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)

library(tidyverse)
```

## A Graph is a Graph is a Graph

-   Most visualisations will boil down to:

```{r, echo = FALSE}
x <- c(1, 3, 4, 5, 6)
y <- c(6, 3, 5, 1, 2)
plot(x, y, type = "l")

```

## Building Blocks of a Graph

<center>

```{r, echo = FALSE, comment = NA}
table <- data.frame(as.table(setNames(x, y)))
colnames(table)[1] <- "x"
colnames(table)[2] <- "y"
print(table, row.names = FALSE)
```

<center>

## Building Blocks of a Graph

```{r, echo = FALSE}
plot(x, y, type = "n")
```

## Building Blocks of a Graph

```{r, echo = FALSE}
plot(x, y, type = "l", axes = FALSE)
```

## Why?

> -   Understanding breakdown of your graph is important:

> -   This is core of ggplot2!

> -   Will make plotting of perfect graph easier

> -   Process of adding layers

## Explore data

```{r, include = TRUE, echo = TRUE, message = FALSE}
colnames(diamonds)
```

::: notes
dataset diamonds is one we work with for this example
:::

## Explore data

```{r, include = TRUE, echo = TRUE, message = FALSE, comment = NA}
summary(diamonds)
```

## Setup

Tell ggplot what data set to use

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds)
```

## Adding axis

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds, aes (x = carat, y = price))
```

## Adding layers

```{r, include = TRUE, echo = TRUE, message = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

## What is happening:

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

> -   ggplot(diamonds) loads data frame
> -   plus (+) tells ggplot() that there is more to add
> -   geo_point() defines type of plot
> -   aes(x = , y =) defines the variables

## Change colour

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point()
```

::: notes
Automatically adds legend
:::

## Adding more layers

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

**Geom_smooth() adds a trend line over an existing ggplot2 visualization. By default the trend line that's added is a LOESS smooth line.**

## Alternative coding

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds)+
  geom_point(aes(x = carat, y = price, colour = cut)) +
  geom_smooth(aes(x = carat, y = price, colour = cut))

```

## Change geom_smooth

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE}
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, colour = cut)) +
  geom_smooth(aes(x = carat, y = price))
```

## Change geom_smooth

```{r, include = TRUE, echo = TRUE, message = FALSE}
# Try changing x to either "x", "y", "z", "depth" and y to "x", "y", "z", "depth"
ggplot(diamonds) +
  geom_point(aes(x = , y = , colour = cut)) +
  geom_smooth(aes(x = , y = ))
```

## Change shape

```{r, include = TRUE, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE}
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, colour = cut, shape = cut)) +
  geom_smooth(aes(x = carat, y = price))
```

## Change shape

```{r, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE}
#Try using shape = color
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, colour = cut, shape = color)) +
  geom_smooth(aes(x = carat, y = price))
```

## Your turn!

-   Use mpg data set
-   Explore the column names
-   Explore data with base R plot
-   Scatterplot of engine size (displ) & fuel efficiency (hwy)
-   Colour by type of car (class)
-   Add an average line

## Result column names

```{r, include = TRUE, echo= TRUE, message = FALSE, warning = FALSE}
colnames(mpg)
```

## Result histogram {.smaller .columns-2}

```{r, out.width = '500px', include = TRUE, echo= TRUE}
hist(mpg$displ)
```

```{r, out.width = '500px', include = TRUE, echo= TRUE}
hist(mpg$hwy)
```

## Result scatterplot

```{r, include = TRUE, echo= TRUE, message = FALSE, warning = FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) +
  geom_smooth(aes(x = displ, y = hwy))
```

## Next R cafe

-   More ggplot
-   Make it pretty:
    -   Axis
    -   Legend
    -   Title
-   Other graph types

## Questions or Suggestions?

Email us:\

Dr Priyanka Baloni: [pbaloni\@purdue.edu](mailto:pbaloni@purdue.edu){.email}\

Dr Anke Tukker: [atukker\@purdue.edu](mailto:atukker@purdue.edu){.email}
